Global patterns of forest autotrophic carbon fluxes

### 2020-08-06 manuscript/manuscript draft.Rmd
working on discusion
(still a mess, pushing now to avoid conflicted copies)

Forest C fluxes decrease with latitude [e.g., @luyssaert_co_2007; @piao_forest_2010; @gillman_latitude_2015; @li_mapping_2019], but it remains unclear whether and how the shape of this relationship varies among fluxes. 
For instance, gross primary productivity ($GPP$) increases continuously with decreasing latitude and is indisputably highest in tropical forests [@luyssaert_co_2007; @beer_terrestrial_2010; @jung_global_2011; @badgley_terrestrial_2019; @li_mapping_2019]. 
In contrast, some studies have suggested that net primary productivity ($NPP$), or its aboveground portion ($ANPP$), exhibits a less distinct increase from temperate to tropical forests [@luyssaert_co_2007]--or even a decrease [@huston_global_2009, but see @gillman_latitude_2015]. 
A shallower increase in $NPP$ than in $GPP$ with decreasing latitude would align with the suggestion that tropical forests tend to have low carbon use efficiency [$CUE$= $NPP$/$GPP$; @de_lucia_forest_2007; @malhi_productivity_2012; @anderson-teixeira_carbon_2016].
However, accurately comparing relationships of multiple forest C fluxes across latitude requires a large database with standardization for methodology, stand history, and management, which to date has not been possible.

@@

Our analysis of a large global database (ForC) reveals how autotrophic carbon fluxes in mature forests vary with latitude and climate on a global scale. We show that, across all nine forest autotrophic C fluxes analyzed, C cycling decreases continually with latitude (*H1.1*; Fig. 2)--a finding that confirms multiple previous studies but contradicts the idea that productivity of temperate forests rivals that of tropical forests [@huston_global_2009]. C fluxes generally increase in proportion to one another (*H1.2*), with few differences in allocation detectable at this global scale (Fig. S2) and with component fluxes summing appropriately to larger fluxes (Fig. 2), indicating no major, systematic omissions or overestimations of flux components. However, climate explained lower proportions of variability among subsidiary C fluxes (*e.g.*, $ANPP_{stem}$, $BNPP_{fine.root}$, $R_{root}$; Fig. 2; Table S2). 
Latitudinal variation in C fluxes is primarily attributable to temperature-related variables (*H3, H4*), particularly MAT (Figs. 3-4). Water availability is also influential, but generally of secondary importance across the climate space occupied by forests (Figs. 3-4). Temperature seasonality and growing season length are closely correlated with MAT and are strong predictors of C fluxes (*H4*; Figs. 4e-f, S2, S6-S7), though growing season length did not improve upon MAT as a predictor. Within the growing season, the influence of climate on C cycling is smaller but still significant for a number of C fluxes (*H5*; Fig. S9; Table S4).

These findings clarify how forest C fluxes vary with latitude and climate on a global scale. Past studies have differed in their conclusions regarding the relationship between C fluxes and latitude or its correlates (Table 1, *H1*)--quite possibly because of lack of standardization with respect to stand age and disturbance history. Our findings indicate that, among mature, undisturbed stands, forest C fluxes are unambiguously highest in the tropical regions, and the relationship with latitude is approximately linear (Fig. 2). This contrasts with the suggestion that productivity of temperate forests is similar to that of tropical forests [@huston_global_2009]. Compared to tropical forests, the temperate forest biome has experienced more widespread anthropogenic disturbance and has a larger fraction of secondary stands [@potapov_mapping_2008; @poulter_global_2018], so analyses comparing across latitudinal gradients without controlling for stand age risk confounding age with biome effects. In addition, because carbon allocation varies with stand age [@de_lucia_forest_2007; @anderson-teixeira_altered_2013; @doughty_what_2018], age differences may introduce systematic biases into analyses of C fluxes across latitude or global climatic gradients. For example, woody productivity tends to be higher in rapidly aggrading secondary stands than in old-growth forests, where proportionally more C is allocated to respiration [@de_lucia_forest_2007; @piao_forest_2010; @doughty_what_2018; @kunert_understanding_2019].

We show that C fluxes are broadly consistent in their responses to climate drivers on the global scale, with at most modest trends in C allocation among the variable pairs tested (Figs. 2, S3). 
This parallels the observation that C allocation across multiple C fluxes varies little with respect to climate along a steep tropical elevational gradient [@malhi_variation_2017; but see @moser_elevation_2011], 
and is not surprising given that trees face universal constraints in carbon allocation (**REFS**). 
The one trend in allocation that emerged from our analysis was a tendency for greater C allocation aboveground--and to foliage in particular--in warm tropical climates (Fig. S3). This is consistent with observations of increasing allocation to roots with declining temperature across a tropical elevational gradient [@moser_elevation_2011], 
and with observations and theory predicting relatively higher belowground carbon allocation at higher latitudes [@gill_belowground_2016]. **Krista -- I took out the ANPP:NPP graphs from the SI after looking back at what Helene said in issue #77 -- let me know if you think it's worth putting them back in wrt this point**
It is also consistent with findings that as forest productivity increases, proportionaly more carbon is allocated to $ANPP_{foliage}$ relative to $BNPP_{fine.root}$ [@chen_allometry_2019] or $ANPP_{stem}$ [@hofhansl_new_2015]. **( check / comment on  @litton_carbon_2007 )**

@@

**update this paragraph ([issue 78](https://github.com/forc-db/Global_Productivity/issues/78))** 
One interesting observation was that climate tends to explain more variation in the major fluxes ($GPP$, $NPP$, $R_{auto}$) than in subsidiary fluxes ($BNPP_{fine.root}$, $R_{root}$, $ANPP_{stem}$) as quantified by R^2^ (Fig. 2; Table S2). There are two, non-exclusive, potential explanations for this. First, it may be that methodological variation is larger relative to flux magnitude for some of the subsidiary fluxes. Belowground fluxes in particular are difficult to quantify, and measurement methods for the belowground fluxes considered here may use fundamentally different approaches in different sites (*e.g.*, minirhizotrons, ingrowth cores, or sequential coring for $BNPP_{fine.root}$; root exclusion, stable isotope tracking, or gas exchange of excised roots for $R_{root}$), and sampling depth is variable and often insufficient to capture the full soil profile. $ANPP_{stem}$, which is also poorly explained by latitude or climate, is more straightforward to measure but is subject to variability introduced by differences such as biomass allometries applied and minimum plant size sampled [@clark_measuring_2001]. However, methodological variation and uncertainty affect all of fluxes considered here, and some of the larger fluxes that vary more strongly with respect to climate ($ANPP$, $NPP$) are estimated by summing uncertain component fluxes. Second, differences among variables in the proportion of variation explained by climate may be attributable to more direct climatic control over $GPP$ than subsidiary fluxes. That is, subsidiary fluxes may be shaped by climate both indirectly through its influence on $GPP$ and respiration and directly through any climatic influence on C allocation, as well as many other local- and regional-scale factors (**REFS**).

The latitudinal gradient in forest C flux (Fig. 2) is driven primarily by temperature-related climate variables, the effects of which are moderated by moisture availability (Table 1, *H2-H3*; Figs. 3-4).  MAT and MAP have long been identified as primary global-scale drivers of C fluxes [@lieth_primary_1973; **REFS**; @taylor_temperature_2017]. It is not appropriate to attempt to identify individual mean annual climate variables as mechanistic drivers of C fluxes because many climate variables co-vary across the latitudinal gradient (Fig. S2), because climatic drivers affect forest carbon flux on much shorter time scales than can be captured by annual climate summary variables, and because both climatic conditions and C flux vary intra- and inter-annually around the long-term means. However, it remains informative to consider these relationships. Among the temperature-related climate variables, $MAT$ is generally the most strongly correlated with C fluxes (Table S2)--perhaps in part because site-specific MAT is recorded for the majority of sites in ForC, whereas other variables were extracted from global gridded data products, introducing spatial scaling errors.

>>>>

Forest C fluxes decrease with latitude [e.g., @luyssaert_co_2007; @piao_forest_2010; @gillman_latitude_2015; @li_mapping_2019], but studies have differed in their conclusions regarding the shape of this relationship - quite possibly because of lack of standardization with respect to methodology and stand history.
For instance, studies agree that gross primary productivity ($GPP$) increases continuously with decreasing latitude and is indisputably highest in tropical forests [@luyssaert_co_2007; @beer_terrestrial_2010; @jung_global_2011; @badgley_terrestrial_2019; @li_mapping_2019]. 
In contrast, some studies have suggested that net primary productivity ($NPP$), or its aboveground portion ($ANPP$), exhibits a less distinct increase from temperate to tropical forests [@luyssaert_co_2007]--or even a decrease [@huston_global_2009, but see @gillman_latitude_2015]. 
A shallower increase in $NPP$ than in $GPP$ with decreasing latitude would align with the suggestion that tropical forests tend to have low carbon use efficiency [$CUE$= $NPP$/$GPP$; @de_lucia_forest_2007; @malhi_productivity_2012; @anderson-teixeira_carbon_2016].
Such differences among C fluxes their relationship to latitude could have profound implications for our understanding of the C cycle and its climate sensitivity. However, until recently the potential to compare latitudinal trends across C fluxes has been limited by lack of a large database with standardization for methodology, stand history, and management [@anderson-teixeira_forc_2018].

@@

**Our analysis of a large global database (ForC) clarifies how autotrophic C fluxes in mature forests vary with latitude and climate on a global scale.** 
We show that, across all nine variables analyzed, annual C flux decreases continually with latitude (Fig. 2)--a finding that confirms multiple previous studies but contradicts the idea that productivity of temperate forests rivals or even exceeds that of tropical forests [@luyssaert_co_2007; @huston_global_2009]. 
At this global scale, C fluxes increase approximately in proportion to one another, with component fluxes summing appropriately to larger fluxes and no detectable differences in allocation across latitude or climates (Figs. 2, 4, S3). 
Similarly, we show broad - *albeit* not complete - consistency of climate responses across C fluxes, with the observed latitudinal variation primarily attributable to temperature and its seasonality (Figs. 3-4).
Water availability is also influential, but of secondary importance across the climate space occupied by forests (Figs. 3-4). 
Contrary to prior suggestions that the majority of variation in C cycling is driven by the length of the growing season (**REFS**), we find modest explanatory power of growing season length and small but sometimes significant influence of climate within the growing season (Figs. 4f,S6-S9).
Together, these findings yield a unified understanding of climate's influence on forest C cycling.

**Our findings indicate that, among mature, undisturbed stands, forest C fluxes are unambiguously highest in the tropical regions, and the relationship with both latitude and $MAT$ is approximately linear (Figs. 2, 4).** 
This contrasts with the suggestion that C fluxes (e.g., $NPP$, $ANPP$, $ANPP_{stem}$) of temperate forests are similar to or even greater than that of tropical forests [@luyssaert_co_2007; @huston_global_2009]. 
Previous indications of such a pattern may have been an artifact of differences in stand age across biomes.
Compared to tropical forests, the temperate forest biome has experienced more widespread anthropogenic disturbance and has a larger fraction of secondary stands [@potapov_mapping_2008; @poulter_global_2018], so analyses comparing across latitudinal gradients without controlling for stand age risk confounding age with biome effects. 
Because carbon allocation varies with stand age [@de_lucia_forest_2007; @anderson-teixeira_altered_2013; @doughty_what_2018], age differences may introduce systematic biases into analyses of C fluxes across latitude or global climatic gradients. 
For example, woody productivity tends to be higher in rapidly aggrading secondary stands than in old-growth forests, where proportionally more C is allocated to respiration and non-woody productivity [@de_lucia_forest_2007; @piao_forest_2010; @doughty_what_2018; @kunert_understanding_2019].
Thus, findings that temperate forest productivity rivals that of tropical forests are likely an artifact of different forest ages across biomes.

**We show that C fluxes are broadly consistent in their responses to climate drivers on the global scale, with at most modest trends in C allocation among the variable pairs tested (Figs. 2, S3).** 
This parallels the observation that C allocation across multiple C fluxes varies little with respect to climate along a steep tropical elevational gradient [@malhi_variation_2017; but see @moser_elevation_2011], 
and is not surprising given that trees face universal constraints in carbon allocation (**REFS**). 
The one trend in allocation that emerged from our analysis was a tendency for greater C allocation aboveground--and to foliage in particular--in warm tropical climates (Fig. S3). This is consistent with observations of increasing allocation to roots with declining temperature across a tropical elevational gradient [@moser_elevation_2011], 
and with observations and theory predicting relatively higher belowground carbon allocation at higher latitudes [@gill_belowground_2016]. **Krista -- I took out the ANPP:NPP graphs from the SI after looking back at what Helene said in issue #77 -- let me know if you think it's worth putting them back in wrt this point**
It is also consistent with findings that as forest productivity increases, proportionally more carbon is allocated to $ANPP_{foliage}$ relative to $BNPP_{fine.root}$ [@chen_allometry_2019] or $ANPP_{stem}$ [@hofhansl_new_2015]. **( check / comment on  @litton_carbon_2007 )**

@@

**update this paragraph ([issue 78](https://github.com/forc-db/Global_Productivity/issues/78))** 
*Despite the broad consistency of climate responses across C fluxes, climate explains lower proportions of variability among some of the subsidiary C fluxes (*e.g.*, $ANPP_{stem}$, $BNPP$  $BNPP_{fine.root)$; Fig. 2; Tables S2, S6).* 
There are two, non-exclusive, potential explanations for this. 
First, it may be that methodological variation is larger relative to flux magnitude for some of the subsidiary fluxes. 
Belowground fluxes in particular are difficult to quantify, and measurement methods for the belowground fluxes considered here may use fundamentally different approaches in different sites (*e.g.*, minirhizotrons, ingrowth cores, or sequential coring for $BNPP_{fine.root}$; root exclusion, stable isotope tracking, or gas exchange of excised roots for $R_{root}$), and sampling depth is variable and often insufficient to capture the full soil profile. 
$ANPP_{stem}$, which is also poorly explained by latitude or climate, is more straightforward to measure but is subject to variability introduced by differences such as biomass allometries applied and minimum plant size sampled [@clark_measuring_2001]. 
However, methodological variation and uncertainty affect all of fluxes considered here, and some of the larger fluxes that vary more strongly with respect to climate ($ANPP$, $NPP$) are estimated by summing uncertain component fluxes. 
Second, differences among variables in the proportion of variation explained by climate may be attributable to more direct climatic control over $GPP$ than subsidiary fluxes. 
That is, subsidiary fluxes may be shaped by climate both indirectly through its influence on $GPP$ and respiration and directly through any climatic influence on C allocation, as well as many other local- and regional-scale factors (**REFS**).

*At least at this global scale, we find little evidence of any non-linearity in temperature's influence on C flux.* 
The latitudinal gradient in forest C flux (Fig. 2) is driven primarily by temperature-related climate variables, the effects of which are moderated by moisture availability (Table 1, *H2-H3*; Figs. 3-4).  MAT and MAP have long been identified as primary global-scale drivers of C fluxes [@lieth_primary_1973; **REFS**; @taylor_temperature_2017]. It is not appropriate to attempt to identify individual mean annual climate variables as mechanistic drivers of C fluxes because many climate variables co-vary across the latitudinal gradient (Fig. S2), because climatic drivers affect forest carbon flux on much shorter time scales than can be captured by annual climate summary variables, and because both climatic conditions and C flux vary intra- and inter-annually around the long-term means. However, it remains informative to consider these relationships. Among the temperature-related climate variables, $MAT$ is generally the most strongly correlated with C fluxes (Table S2)--perhaps in part because site-specific MAT is recorded for the majority of sites in ForC, whereas other variables were extracted from global gridded data products, introducing spatial scaling errors.
